Commutativity of operation-based CRDTs \\ CIS 673 Project Report

### 2020-04-15 report/draft.tex
Sections 3 and 4

\title{CIS 673 Project Report}
\author{Gautam Mohan and Konstantinos Kallas}

@@

\section{Conflict Free Replicated Data Types}

\TODO{Fill in with some background information on CRDTs}~\cite{shapiro2011conflict}

\TODO{Say something about both State and Op-based CRDTs.}

\TODO{State correctness conditions Op-based only}

\TODO{Say why we only need to prove commut of effects}

@@

We now describe how to encode the update methods. Each update method
has two parts, a \texttt{prepare} and an \texttt{effect}. In Servois
we only encode effects as methods, as we care about how concurrent
updates (that happened in different replicas) can affect the state of
another replica. Intuitively, we want to show that the effects $e_1$,
$e_2$ of two concurrent updates $u_1$, $u_2$ that happened in replicas
$r_1$ and $r_2$ will commute in a third replica $r_3$. In order to
correctly do that, we need to derive preconditions for all the effect
methods by using their prepare methods. More precisely, the
precondition on the arguments of an effect $e$ of an update $u$ that
happened in a replica with state $s'$ must be equivalent to the
postcondition of $u$'s prepare method after it has been applied
$s'$. \TODO{Is that valid?}

\TODO{Figure out what we have to do with effect preconditions. Maybe
  we need to encode them with ite to make sure that if they fail, it
  means that an update is not enabled when it should be?}

@@

\TODO{Briefly mention the threats to validity, the fact that we derive
  preconditions based on prepare methods but we don't check them, and
  the fact that we manually encode the postconditions of each method.}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.5\textwidth]{grow-only-set-definition}
    \caption{Grow-only Set}
    \label{fig:gset-def}
\end{figure}

>>>>

\title{Commutativity of operation-based CRDTs \\ CIS 673 Project Report}
\author{Gautam Mohan and Konstantinos Kallas}

@@

\section{Conflict-Free Replicated Data Types}

Conflict-free replicated data types are objects that can be
implemented on top of eventually consistent datastores guaranteeing
strong convergence without any need for conflict resolution or
roll-back. Conflict-freedom ensures safety and liveness despite any
number of failures. This section contains necessary background
information for CRDTs that is adapted from~\cite{shapiro2011conflict}.

The standard correctness condition for conflict-free replicated data
types is strong eventual consistency (SEC), which is slightly stronger
than eventual consistency. Its main difference is strong convergence,
namely that replicas that have delivered the same updates have
equivalent states. Objects that are SEC can be implemented on top of
replicated data stores that do not synchronize, allowing them to be
always available for both reads and writes independently of network
conditions.

Conflict-free replicated data types can be classified in two types,
the state-based ones and the operation-based ones. The intuitive
difference is that in state-based CRDTs replicas merge their states
periodically, while in operation-based CRDTs all replicas broadcast
all of their updates to all other replicas. Both types of CRDTs can be
proven to be SEC given some conditions on their specifications
hold. In this project we focus on operation-based CRDTs, since their
correctness, i.e. satisfying strong eventual consistency, is implied
by their commutativity of their operations.

\subsection{Operation-based CRDTs}

An op-based object can be specified by the following tuple $(S, s^0,
Q, U)$, where $S$ is the state domain, $s^0$ is the initial state, $Q$
is a set of query methods, and $U$ are update methods. Update methods
are split into a pair $(t,u)$, where $t$ is a side-effect-free
\texttt{prepare} method and $u$ is an \texttt{effect} method (the
arguments of which may differ. Both \texttt{prepare} and
\texttt{effect} methods have enabling preconditions. The
\texttt{prepare} method is executed at a single replica where the
operation is invoked, generating the arguments for the \texttt{effect}
without modifying the state. The \texttt{effect} is executed
immediately after it in that replica, and it is then broadcast to all
other replicas for execution.

An important assumption is that the underlying broadcast communication
protocol delivers messages preserving causal order. This means that if
a \texttt{prepare} $t_j$ is executed on a replica after it has
received an effect $e_i$, then all other replicas will receive $e_j$
after they receive $e_i$. Concurrent updates (i.e. ones that happened
in different source replicas can be delivered in any order.

Given these assumptions, they prove in
Theorem~2.2~\cite{shapiro2011conflict} that the commutativity of all
concurrent effect operations of an op-based CRDT is a sufficient
condition for convergence. Note that they make one more assumption,
namely that causally-ordered broadcast is sufficient to ensure that an
effect is enabled in all other replicas if it was enabled in the
initial one.

%% Effect precondition may require that an effect is delayed, so in order
%% to prove convergence, one would need to prove liveness, that
%% eventually an effects precondition is satisfied. They only focus on
%% effect preconditions that are satisfied by the delivery order in that
%% paper, and that is what we focus on too.

@@

We now describe how to encode the update methods. Each update method
has two parts, a \texttt{prepare} and an \texttt{effect}. In Servois
we only encode effects as methods, as we are only interested in the
commutativity of concurrent updates.
%% \TODO{Briefly mention that we don't care about commutativity of
%%   prepare-effect with effect because of the underlying causal
%%   delivery protocol.}
Intuitively, we want to show that the effects $e_1$, $e_2$ of two
concurrent updates $u_1$, $u_2$ that happened in replicas $r_1$ and
$r_2$ will commute in a third replica $r_3$. In order to correctly do
that, we need to derive preconditions for all the effect methods by
using their prepare methods. More precisely, the precondition on the
arguments of an effect $e$ of an update $u$ that happened in a replica
with state $s'$ must be equivalent to the postcondition of $u$'s
prepare method after it has been applied $s'$.
%% \TODO{Is that valid?}

\TODO{(Maybe -- If we have time) Figure out what we have to do with
  effect preconditions. Maybe we need to encode them with ite to make
  sure that if they fail, it means that an update is not enabled when
  it should be?}

@@

\paragraph{Threats to validity:}

We would like to briefly mention two threats to the validity of our
methodology. First, both \texttt{effect} pre and postconditions are
derived from the specification manually, introducing a possible point
of error. Second, we do not check whether the delivery preconditions
of \texttt{effect} methods are indeed satisfied by the causal delivery
order (similarly to~\cite{shapiro2011conflict}).

%% \TODO{Briefly mention the threats to validity, the fact that we derive
%%   preconditions based on prepare methods but we don't check them, and
%%   the fact that we manually encode the postconditions of each
%%   method. Also that we don't check whether delivery preconditions (of
%%   effects) are indeed satisfied by the causal delivery order.}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.4\textwidth]{grow-only-set-definition}
    \caption{Grow-only Set}
    \label{fig:gset-def}
\end{figure}
